---
title: "Diamond Prices Analysis"
author: "TJ Cook"
date: "`3/16/26"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)

library(ggplot2)
library(dplyr)
library(knitr)
```

---

# Diamond Prices

This analysis explores the relationship between diamond weight (carat) and price using the `ggplot2::diamonds` dataset.

## Question 1 — Filter & Count Table

We retain only diamonds whose c**olor** is D (best), G (middle-grade), or J (worst), and whose **clarity** is IF (best), VS1 (middle-grade), or I1 (worst).

```{r q1-filter}
diamonds_sub <- diamonds %>% filter(color %in% c("D", "G", "J"), clarity %in% c("I1", "VS1", "IF")) %>% mutate(color   = droplevels(color), clarity = droplevels(clarity))
```

```{r q1-table}
diamonds_sub %>% count(color, clarity) %>% tidyr::pivot_wider(names_from = clarity, values_from = n, values_fill = 0) %>% kable()
```

The filtered dataset contains `r nrow(diamonds_sub)` diamonds spread across
nine color–clarity combinations. IF diamonds are considerably rarer than VS1 or
I1 diamonds across all three color grades. [Question 1 complete]

## Question 2 — Price vs. Carat (Overall)

```{r q2-scatter, fig.width=7, fig.height=5}
ggplot(diamonds_sub, aes(x = carat, y = price)) + geom_point(alpha = 0.5) + geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 1) + theme_minimal()
```

There is a clear **positive relationship** between carat weight and price —
heavier diamonds are substantially more expensive. [Question 2 complete]

## Question 3 — Price vs. Carat by Color

```{r q3-color, fig.width=7, fig.height=5}
ggplot(diamonds_sub, aes(x = carat, y = price, color = color)) + geom_point(alpha = 0.5) + geom_smooth(method = "lm", se = FALSE, linewidth = 1) + scale_color_brewer(palette = "Dark2", name    = "Color\n(D = best, J = worst)") + theme_minimal()
```

All three color grades show the same positive price–carat trend, but there are differences. **D-color** (best) diamonds trend line sits highest. **G-color** diamonds fall in the middle, and **J-color** (worst) diamonds fetch the lowest prices per carat. [Question 3 complete]

## Question 4 — Price vs. Carat by Color & Clarity

```{r q4-color-clarity, fig.width=8, fig.height=5}
ggplot(diamonds_sub, aes(x = carat, y = price, color    = color, shape    = clarity, linetype = clarity)) + geom_point(alpha = 0.5) + geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) + scale_color_brewer(palette = "Dark2", name    = "Color\n(D = best, J = worst)") + theme_minimal()
```

Mapping clarity to both point shape and line type results in a **difficult-to-read** plot. [Question 4 complete]

## Question 5 — Faceted Grid: Color × Clarity

```{r q5-facet, fig.width=9, fig.height=7}
ggplot(diamonds_sub, aes(x = carat, y = price)) + geom_point(alpha = 0.4, size = 0.8) + geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 0.9) + facet_grid(color ~ clarity, labeller = labeller(color = c(D = "Color D (best)", G = "Color G (mid)", J = "Color J (worst)"), clarity = c(I1  = "Clarity I1 (worst)", VS1 = "Clarity VS1 (mid)", IF  = "Clarity IF (best)")))  + theme_minimal(base_size = 10) + theme(strip.text = element_text(face = "bold"))
```

All panels show a positive price–carat relationship. The **D–IF** combination (top-right panel) has the steepest slope and the fewest data points. Conversely, **J–I1** (bottom-left) has the shallowest slope. In general, moving toward better color *and* better clarity steepens the price-per-carat. [Question 5 complete]

## Question 6 — Average Price per Carat by Color & Clarity

```{r q6-ratio}
diamonds_ppc <- diamonds_sub %>% mutate(price_per_carat = price / carat)

ppc_summary <- diamonds_ppc %>% group_by(color, clarity) %>% summarize(avg_price_per_carat = round(mean(price_per_carat), 2), n = n(), .groups = "drop") %>% arrange(desc(avg_price_per_carat))

kable(ppc_summary, col.names = c("Color", "Clarity", "Avg Price per Carat (USD)", "N"))
```

The table largely agrees with the visual impressions from Questions 3 and 5.
**D–IF** (best color, best clarity) tops the ranking with the highest average
price per carat, confirming what the facet plot suggested. J–I1 (worst color,
worst clarity) sits near the bottom. One noteworthy quirk: within a given color
grade, IF clarity does not always rank above VS1 in price-per-carat — this can
happen because the IF cells have very few observations and may skew toward
smaller stones that don't yet command their full premium. Overall, the numeric
ranking is consistent with the graphical story: better quality diamonds fetch
more per carat.^[Question 6 complete]
```
